Benefits and risks associated with children’s and adolescents’ interactions with electronic screens: An umbrella review

### 2023-03-12 reports/manuscript.Rmd
Address current coauthor comments

Citing the negative effects of screens on health (e.g., increased risk of obesity) and health-related behaviours (e.g., sleep), guidelines from the World Health Organisation [@whoGuidelinesPhysicalActivity2019] and numerous government agencies [@australiangovernmentPhysicalActivityExercise2021;@Canadian24HourMovement2016] and statements by expert groups [@AAPMediaUseSchoolAged2016] have recommended that young people's time spent using electronic media devices for entertainment purposes should be limited.
For example, the Australian Government guidelines regarding sedentary behaviour recommend that young children (under the age of two) should not spend any time watching screens.
They also recommend that children aged 2-5 years should spend no more than one hour engaged in recreational sedentary screen use per day, while children aged 5-12 and adolescents should spend no more than two hours.
However, recent evidence suggests that longer exposures may not have adverse effects on children's behaviour or mental health---and might, in fact, benefit their well-being---as long as exposure does not reach extreme levels (e.g., 7 hours per day)[@fergusonEverythingModerationModerate2017].
Some research also indicates that content (e.g., video games vs television programs) plays an important role in determining the potential benefit or harm of youths' exposure to screen-based media [@przybylskiLargeScaleTestGoldilocks2017].
Indeed, educational screen time is positively related to educational outcomes [@sandersTypeScreenTime2019].
This evidence has led some researchers to argue that a more nuanced approach to screen time guidelines is required [@kayeConceptualMethodologicalMayhem2020].

In 2016, the American Academy of Pediatrics used a narrative review to examine the benefits and risks of children and adolescents' electronic media [@chassiakosChildrenAdolescentsDigital2016] as a basis for updating their guidelines about screen use [@AAPMediaUseSchoolAged2016].
Since then, a large number of systematic reviews and meta-analyses have provided evidence about the potential benefits and risks of screen use.
<span id = "r2_9">
While there have been other overviews of reviews on screen time, these have tended to focus on a single domain (e.g., health [@stiglicEffectsScreentimeHealth2019]), focus on a particular exposure (e.g., social media [@valkenburgSocialMediaUse2022;@arias-delatorreRelationshipDepressionUse2020]) or provide only a narrative summary of the literature [@orbenTeenagersScreensSocial2020].
<span id = "r1_2">
Focusing on a single domain or exposure makes it difficult to understand what trade-offs are involved in any guidelines around screen use.
For example, prohibiting screen use might reduce exposure to advertising but may also thwart learning opportunities from interactive educational tools. Reviews on either of these exposures or outcomes would likely miss being able to quantify these trade-offs.
Overviews are one method of evidence synthesis that helps address these trade-offs, by providing 'user-friendly' summaries of a field of research [@pollockChapterOverviewsReviews2022].
These overviews provide a reference point for the field and allow for easier comparison of risks and benefits for the same behaviour.</span></span>

In order to synthesise the evidence and support further evidence-based guideline development and refinement, we reviewed published meta-analyses examining the effects of screen use on children and youth.
<span id = "r2_6">This review synthesises evidence on any outcome of electronic media exposure.
We deliberately do not pre-specify outcomes, in order to get a comprehensive list of areas where there is meta-analytical evidence.
Adopting this broad approach allowed us to provide a holistic perspective on the influence of screens on children's lives.
By synthesising across life domains (e.g., school and home), this review provides evidence to inform guidelines and advice for parents, teachers, pediatricians and other professionals in order to maximise human functioning.</span>

@@

We initially planned to describe the findings of systematic reviews that did not include meta-analyses in a narrative summary alongside the main findings from meta-analytic studies.
However, we determined that combining results from the meta-analyses allowed readers to compare relative strength of associations more easily.
Readers interested in the relevant systematic reviews (i.e., without meta-analysis) can consult the list of references in Supplementary File 4.

@@

Broadly, our findings align with the recommendations of others who suggest that current guidelines may be too simplistic, mischaracterise the strength of the evidence, or do not acknowledge the important nuances of the issue [@elsonPolicyStatementsMedia2019;@ashtonScreenTimeChildren2019;@royalcollegeofpaediatricsandchildhealthHealthImpactsScreen2019].
Our findings suggest that screen use is a complex issue, with associations based not just on duration and device type, but also on the content and the environment in which the exposure occurs.
Many current guidelines simplify this complex relationship as something that should be minimised in all instances [@australiangovernmentPhysicalActivityExercise2021;@whoGuidelinesPhysicalActivity2019].
We suggest that future guidelines need to embrace the complexity of the issue, to give parents and clinicians specific information to weigh the pros and cons of interactions with screens.

@@

Our high-level approach limits the feasibility of examining fine-grained details of the individual studies.
For example, we did not examine moderators beyond age, nor did we rate the risk of bias for the individual studies.
Thus, our assessment of evidence quality was restricted to statistical credibility, rather than a more complete assessment of quality (e.g., GRADE [@guyattGRADEGuidelinesIntroduction2011]).
As such, we made decisions regarding the credibility of evidence, where others may have used different thresholds or metrics.
For this reason, we provide the complete results in the supplementary material, along with the dataset for others to consider alternative criteria.

<span id="r1_4">Our high-level approach also means that we could not engage with the specific mechanisms behind each association, and as such, we cannot make strong claims on the directions of causality.
These likely depend on the specific exposure and outcome.
It is tempting for people to draw inferences that the associations are due to screen time causing these outcomes, but we cannot rule out reverse causality, a third variable, or some combination of influences.
Many of the individual reviews go into more detail about the strength of the evidence for causal associations, but those judgements were difficult to synthesise across more than 200 reviews.
Readers who wish to more deeply understand one specific relationship are directed to the cited review for that effect, where the authors could engage more deeply with the mechanisms.</span>
<span id="r2_16">We converted all effect sizes to a common metric (Pearson's r) to allow for comparisons of magnitude, but acknowledge that this assumes a linear relationship.
Some previous research suggests that associations are typically linear [@sandersTypeScreenTime2019].
However, others have identified instances where non-linear relationships exist, especially for very high levels of screen time [@twengeMoreTimeTechnology2019; @przybylskiLargeScaleTestGoldilocks2017; @kellySocialMediaUse2018].
Additionally, our conversion may not always adequately account for differences in study design or measures of exposures and outcomes.
Care is needed, therefore, when interpreting the effect sizes.</span>
<span id = "r3_2_2">In addition, reviews provide only historical evidence which may not keep up with the changing ways children can engage with screens.
While our synthesis of the existing evidence provides information about how screens might have influenced children in the past, it is difficult to know if these findings will translate to new forms of technology in the future.</span>

>>>>

Citing the negative effects of screens on health (e.g., increased risk of obesity) and health-related behaviours (e.g., sleep), guidelines from the World Health Organisation [@whoGuidelinesPhysicalActivity2019] and numerous government agencies [@australiangovernmentPhysicalActivityExercise2021;@Canadian24HourMovement2016] and statements by expert groups [@AAPMediaUseSchoolAged2016] have recommended that young people's time spent using electronic media devices for entertainment purposes should be limited.
<span id = "r2_8">For example, the Australian Government guidelines regarding sedentary behaviour recommend that young children (under the age of two) should not spend any time watching screens.
They also recommend that children aged 2-5 years should spend no more than one hour engaged in recreational sedentary screen use per day, while children aged 5-12 and adolescents should spend no more than two hours.
However, recent evidence suggests that longer exposures may not have adverse effects on children's behaviour or mental health---and might, in fact, benefit their well-being---as long as exposure does not reach extreme levels (e.g., 7 hours per day)[@fergusonEverythingModerationModerate2017].
Some research also indicates that content (e.g., video games vs television programs) plays an important role in determining the potential benefit or harm of youths' exposure to screen-based media [@przybylskiLargeScaleTestGoldilocks2017].
Indeed, educational screen time is positively related to educational outcomes [@sandersTypeScreenTime2019].
This evidence has led some researchers to argue that a more nuanced approach to screen time guidelines is required [@kayeConceptualMethodologicalMayhem2020].
</span>

In 2016, the American Academy of Pediatrics used a narrative review to examine the benefits and risks of children and adolescents' electronic media [@chassiakosChildrenAdolescentsDigital2016] as a basis for updating their guidelines about screen use [@AAPMediaUseSchoolAged2016].
Since then, a large number of systematic reviews and meta-analyses have provided evidence about the potential benefits and risks of screen use.
<span id = "r2_9">
While there have been other overviews of reviews on screen time, these have tended to focus on a single domain (e.g., health [@stiglicEffectsScreentimeHealth2019]), focus on a particular exposure (e.g., social media [@valkenburgSocialMediaUse2022;@arias-delatorreRelationshipDepressionUse2020]) or provide only a narrative summary of the literature [@orbenTeenagersScreensSocial2020].
<span id = "r1_2">
Focusing on a single domain or exposure makes it difficult to understand what trade-offs are involved in any guidelines around screen use.
For example, prohibiting screen use might reduce exposure to advertising but may also thwart learning opportunities from interactive educational tools. Reviews on either of these exposures or outcomes would likely miss being able to quantify these trade-offs.
Overviews are one method of evidence synthesis that helps address these trade-offs, by providing 'user-friendly' summaries of a field of research [@pollockChapterOverviewsReviews2022].
These overviews provide a reference point for the field and allow for easier comparison of risks and benefits for the same behaviour.
By analogy, reading is a sedentary behaviour, and only by comparing the health risks against the educational benefits can researchers and policymakers make clear recommendations about what young people should do.</span></span>

In order to synthesise the evidence and support further evidence-based guideline development and refinement, we reviewed published meta-analyses examining the effects of screen use on children and youth.
<span id = "r2_6">This review synthesises evidence on any outcome of electronic media exposure.
We deliberately did not pre-specify outcomes, in order to get a comprehensive list of areas where there is meta-analytical evidence.
Adopting this broad approach allowed us to provide a holistic perspective on the influence of screens on children's lives.
By synthesising across life domains (e.g., school and home), this review provides evidence to inform guidelines and advice for parents, teachers, pediatricians and other professionals in order to maximise human functioning.</span>

@@

<span id = "r2_20">
As described above, we have summarised the meta-analytic findings from all included systematic reviews.
In our protocol, we originally planned to also conduct a narrative synthesis of all systematic reviews, even those without meta-analyses.
However, we determined that combining results from the meta-analyses alone allow readers to compare relative strength of associations more easily.
Readers interested in the relevant systematic reviews (i.e., without meta-analysis) can consult the list of references in Supplementary File 4.
</span>

@@

Broadly, our findings align with the recommendations of others who suggest that current guidelines may be too simplistic, mischaracterise the strength of the evidence, or do not acknowledge the important nuances of the issue [@elsonPolicyStatementsMedia2019;@ashtonScreenTimeChildren2019;@royalcollegeofpaediatricsandchildhealthHealthImpactsScreen2019].
Our findings suggest that screen use is a complex issue, with associations based not just on duration and device type, but also on the content and the environment in which the exposure occurs.
Many current guidelines simplify this complex relationship as something that should be minimised [@australiangovernmentPhysicalActivityExercise2021;@whoGuidelinesPhysicalActivity2019].
We suggest that future guidelines need to embrace the complexity of the issue, to give parents and clinicians specific information to weigh the pros and cons of interactions with screens.

<span id = "r2_7">
In particular, our results support the the continuing trend of guidelines moving away from recommendations to reduce 'screen time', and instead focusing on the type of screen time.
For example, our findings suggest that guidelines should discourage high levels of social media and internet use.
Guidelines may also consider adapting recommendations that promote the use of educational apps and video games, although these recommendations need to be balanced against the (very small) risks to adiposity.
</span>

@@

Our high-level approach limits the feasibility of examining fine-grained details of the individual studies.
For example, we did not examine moderators beyond age, nor did we rate the risk of bias for the individual studies.
Thus, our assessment of evidence quality was restricted to statistical credibility, rather than a more complete assessment of quality (e.g., GRADE [@guyattGRADEGuidelinesIntroduction2011]).
As such, we made decisions regarding the credibility of evidence, where others may have used different thresholds or metrics.
<span id = "r2_24">In addition, when faced with duplicate outcome/exposure combinations we chose to keep the one with the largest pooled sample size, assuming that this would capture the most comprehensive and most recent review.
Inspection of the excluded effect sizes suggests that this decision was not that impactful: our results would have been almost exactly the same has we used the number of included studies (*k*) or the most recent review by publication year.
However, we provide the complete results in the supplementary material, along with the dataset for others to consider alternative criteria.</span>

<span id="r1_4">Our high-level approach also means that we could not engage with the specific mechanisms behind each association, and as such, we cannot make strong claims on the directions of causality.
These likely depend on the specific exposure and outcome.
It is tempting for people to draw inferences that the associations are due to screen time causing these outcomes, but we cannot rule out reverse causality, a third variable, or some combination of influences.
Many of the individual reviews go into more detail about the strength of the evidence for causal associations, but those judgements were difficult to synthesise across more than 200 reviews.
Readers who wish to more deeply understand one specific relationship are directed to the cited review for that effect, where the authors could engage more deeply with the mechanisms.</span>

<span id="r2_16">We converted all effect sizes to a common metric (Pearson's r) to allow for comparisons of magnitude, but acknowledge that this assumes a linear relationship between the variables.
Some previous research suggests that associations are typically linear [@sandersTypeScreenTime2019].
However, others have identified instances where non-linear relationships exist, especially for very high levels of screen time [@twengeMoreTimeTechnology2019; @przybylskiLargeScaleTestGoldilocks2017; @kellySocialMediaUse2018].
Additionally, our conversion may not always adequately account for differences in study design or measures of exposures and outcomes.
Care is needed, therefore, when interpreting the effect sizes.</span>
<span id = "r3_2_2">In addition, reviews provide only historical evidence which may not keep up with the changing ways children can engage with screens.
While our synthesis of the existing evidence provides information about how screens might have influenced children in the past, it is difficult to know if these findings will translate to new forms of technology in the future.</span>

### 2023-03-12 reports/revision_letter.Rmd
Address current coauthor comments

One of the key strengths of our review is that it summarises what is available in the literature, rather than trying to *a-priori* predict what areas should be investigated.
That is, our review provides a comprehensive summary of what meta-analytic evidence exists.
We have added some additional text to this effect in the *Introduction*:

@@

@@

We feel that we have amply addressed the contribution of this review to the field.
Guidelines should reflect both the most recent evidence, and the most comprehensive evidence.
Many of the previous reviews used to justify guidelines focused on narrow ranges of outcomes (e.g., the impact on adiposity).
We think this is a mistake - you cannot make an informed decision about an action if you only examine the risks, you must also examine potential benefits.

@@

We had difficulty understanding the reviewers exact concern here.
We are happy to revise this further if the reviewer does not feel that the other changes made to the introduction improve the set of arguments.

@@

@@

We did not *a-priori* choose which outcomes would be included.
Instead, we chose to summarise the literature on for any meta-analysis where screen exposure was the independent variable.
We have removed the word "plausible", as we agree that makes the sentence unclear and the revised phrasing is more aligned with the eligibility criteria.
Our intention in the introduction is not to preempt the results, which means we do not outline the outcomes found in the introduction.

@@

We agree that "broad approach" does not correctly convey our meaning.
We intended to indicate that in order to make nuanced recommendations, you must have all of the information.
Our approach allowed for a broad set of exposures and outcomes in order to capture all of this information, which is what we intended to convey.
We think our changes elsewhere (see \comment{r1_2} to Reviewer 1, and \comment{r2_6} above) address this.

@@

Per \comment{r2_10} above, we did not pre-specify outcomes and feel it would be inappropriate to revise our inclusion criteria.

@@

We agree this is could be confusing to the reader.
The previous sections outlined the inclusion criteria for including a *review*, whereas these sections focus on reasons for excluding an *effect size* from the review.
We did not require all effect sizes within a review to be eligible or usable for the review to still be included.

@@

We are not clear on what the reviewer means here.
Per the quoted sentence, we intended to include all systematic reviews (not just meta-analyses), but the volume of meta-analyses was such that we felt this was unneeded.
This was a change from our protocol, and thus reported as such.
We are happy to revise this section if the reviewer has a suggestion.

@@

@@

It's possible the reviewer has misunderstood.
In our original plan, we wrote that $P < 0.001$ would be *one of* the criteria for indicating strong evidence (alongside other metrics, such as sample size and heterogeneity).
We realised that this would restrict our 'strong' evidence to only statistically significant effects.
What would mean that we could not include evidence for no effect as strong evidence, which is clearly a mistake.
Our deviation from the protocol (removing the requirement for a P-value threshold) was intended to rectify this.

@@

@@

3. While we agree that this information might be useful for some readers, we found that this made the table difficult to read.
Most reviews presented multiple effect sizes, and once the information to describe each effect was added (e.g., the specifc outcome, exposure, age group, and moderator), the table became overwhelming.
All of the effect sizes (in their original, converted, and recalculated forms) are available in the supplementary material.

@@

1. We agree that the our reporting of the method could have been clearer, and have revised the text in this section to more clearly explain our method.
We disagree with the reviewer that the method was "subjective".
If we had multiple effects for the same exposure/outcome combination, and which had the same sample age restrictions and the same study design restrictions, we chose the one with the largest sample size.
We can certainly see how others might use a different method for eliminating duplicate effects (and we state this in the discussion section), but it is difficult to see how using the sample size could be considered "subjective".

@@

3. We understand the reviewer's concerns regarding the conversions of the effect sizes (although we disagree with the characterisation of our results as "somewhat deceiving").
We chose Pearson's *r* as we wanted to focus not on the precise size of the association, but instead on the *direction* and the *magnitude*.
We think it is worth considering the counterfactual to our approach: had we not converted effect sizes to a common metric readers would likely still make comparisons, but these would be subjective.
We agree that readers should interpret the effect sizes carefully, and have included text to this effect in the *Discussion*:

@@

We are unclear what the reviewer is referring to here, although we think many of the concerns were addressed in \comment{r2_29}.
We feel that a qualitative synthesis of more than 200 effect sizes is outside the scope of this paper.

@@

Even within recreational screen time, our findings do not support guidelines which focus just on minimisation. 
Video games, for example, are a recreational form of screen time which appear to have some benefits for educational and cognitive outcomes.

@@

Our rationale for choosing the NHLBI was that we felt it could be applied equally to all fields than AMSTAR-2.
Reporting standards differ by discipline, and several of the items in AMSTAR-2 are typically only reported in medical fields.
While these should maybe be reported in all fields, we felt that the NHLBI was a more appropriate standard for our review.

>>>>

Screen use is associated with a range of widespread outcomes.
Many reviews focus on a single exposure outcome pairing and miss the trade-offs that are inherent in focusing on just one area (see Reviewer 1, \comment{r1_2} for further discussion).
By focusing on a broad range of outcomes (health and non-health), without trying to *a-priori* predict what areas should be investigated, we simultaneously quantify and compare these trade-offs.

We have added some additional text to this effect in the *Introduction*:

@@

and

@@

Guidelines are important and useful, but they should reflect both the most recent and the most comprehensive evidence.
Many of the previous reviews used to justify guidelines focused on narrow ranges of outcomes (e.g., the impact on adiposity).
We think this is a mistake---you cannot make an informed decision about an action if you only examine the risks, you must weigh these against any potential benefits.
We have included a discussion of this in the *Discussion* section:

`r get_revision("r2_7")`

@@

We appreciate it is not a direct contrast to compare overall limits of screen time against deleterious effects of screen time over 3-4 hours. As a result, we have removed the words 'in contrast':

`r get_revision("r2_8")`

@@

As mentioned in other author responses, we have made a stronger effort to organise the review toward a cohesive purpose.

@@

We did not *a-priori* choose which outcomes would be included.
Instead, we chose to summarise the literature on for any meta-analysis where screen exposure was the independent variable.
We have removed the word "plausible", as we agree that makes the sentence unclear and the revised phrasing is more aligned with the eligibility criteria.
We have tried not to preempt the results section in the introduction, but do now refer to some expected outcomes.
For example:

@@

We agree that "broad approach" does not correctly convey our meaning.
We intended to indicate that in order to make nuanced recommendations, you must have all of the information regarding benefits and risks.
Our approach allowed for a broad set of exposures and outcomes in order to capture all of this information, which is what we intended to convey.
We think our changes elsewhere (see \comment{r1_2} to Reviewer 1, and \comment{r2_6} above) address this.

@@

As per \comment{r2_10} above, we did not pre-specify outcomes and do not want to mislead readers on the process.
We have indicated some anticiapted outcomes, however.

@@

We agree this could be confusing to the reader.
The reason that we had inclusion/exclusion criteria across two sections is that we have separate criteria for reviews, and for the effect sizes within reviews.
The first section outlines the inclusion criteria for including a *review*, whereas the subsequent sections focus on reasons for excluding an *effect size* from the review.
We did not require all effect sizes within a review to be eligible or usable for the review to still be included.
This section now reads:

@@

We agree our phrasing here was unclear. We meant that we intended to narratively summarise systematic reviews, as well as meta-analytic studies. We have changed this phrase to:

`r get_revision("r2_20")`

@@

It is true that harmonising effect sizes does not perfectly retain the original information.
Different measures are able to capture different aspects of the underlying phenomenon.
However, it is better to harmonise for the purposes of crude comparisons than to try and subjectively compare completely different classes of effect sizes. We make this point in the limitations:

In this study we had to choose between subjectively comparing various classes of effect sizes, or attempt to harmonise all effects so quantitative comparisons could easily be made. We elected to harmise the effect sizes so that we could better illustrate differences between studies. However, harmonisation is imperfect, and so to better understand the exact effect from meta-analyses we refer readers back to the original meta-analyses.

@@

We appreciate we may have not explained this clearly.
We have brought our methods into line with other reviews.
In our protocol, we wrote that $P < 0.001$ would be *one of* the criteria for indicating strong evidence (alongside other metrics, such as sample size and heterogeneity).
We realised that this would restrict our 'strong' evidence to only statistically significant effects.
But using this criteria meant that we could never classify non-significant findings as strong evidence regardless of their precision, which is clearly a mistake.
Our deviation from the protocol (removing the requirement for a P-value threshold) was intended to rectify this.

@@

`r get_revision("r2_24")`

@@

3. While we agree that this information might be useful for some readers, we found that this made the table difficult to read.
Most reviews presented multiple effect sizes, and once the information to describe each effect was added (e.g., the specific outcome, exposure, age group, and moderator), the table became overwhelming.
All of the effect sizes (in their original, converted, and recalculated forms) are available in the supplementary material.

@@

1. We agree that our reporting of the method could have been clearer, and have revised the text in this section to more clearly explain our method.
We designed this method to make it as objective as possible.
If we had multiple effects for the same exposure/outcome combination, and which had the same sample age restrictions and the same study design restrictions, we chose the one with the largest sample size.
We can certainly see how others might use a different method for eliminating duplicate effects (and we state this in the discussion section), and we tested whether those methods would have led to different included reviews.

@@

3. We understand the reviewer's concerns regarding the conversions of the effect sizes.
We appreciate that making conversions between effect sizes relies upon some assumptions.
We made the decision what it was better to include these studies rather than omit them, and better to convert the effects rather than presenting different tables for each effect size.
Still, we agree that readers should interpret the effect sizes carefully, and have included text to this effect in the *Discussion*:

@@

We agree that the differences between the effects at different ages are important nuances.
We altered the analytical plan to make it more credible and reliable, but also want to make it so researchers can explore findings that we are less confident in
While we could offer a narrative synthesis of the 200+ effect sizes that did not meet our credibility criteria, we judged this was beyond the scope of this paper.
Doing so would involve some filtering and judgement, exacerbating the selection bias the reviewer points to.
As a result, we have presented those findings in supplementary materials for researchers and practitioners to explore the nuances relevant to them

@@

We agree both guidelines focus on merely reducing recreational screen time and have removed "in all instances".
Even within recreational screen time, however, our findings do not support guidelines which focus just on minimisation. 
Video games, for example, are a recreational form of screen time which appear to have some benefits for educational and cognitive outcomes.

@@

Our rationale for choosing the NHLBI was that we felt it could be applied equally to all fields than AMSTAR-2.
Reporting standards differ by discipline, and several of the items in AMSTAR-2 are typically only reported in medical fields.
AMSTAR-2 and NHLBI overlap on many items, and as AMSTAR-2 taks 20-30min per review we didn't feel that the extra length produced substantially more value.
We felt that the NHLBI was the appropriate standard for our review.
